Using tweet content to detect social media events

### 2020-05-03 Rmd/jse-2019-ms.Rmd
updated ms

---
title: 'Mentored data science research with tweets'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "May 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
---

@@

Twitter has changed the way that we communicate. 
In only 140 characters, users speak their piece on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for text 
analysis now enable instructors to use tweets as pedagogical 
tools in mentoring undergraduate research projects.

@@

@nolan2010computing argue for students to work with real 
data. 
Working with real data allows students to develop skill not 
only in statistical analysis,
but also in related data science skills, including data 
transfer from online sources, data storage, and using data 
from multiple file formats.

@@

We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via topic modeling of tweets and 3) reflections on 
using this data set in research mentoring for undergraduate 
students.

@@

### Collecting tweets

We collected tweets by automated, repeated querying of Twitter's 
streaming API. This involved multiple steps. First, we created a 
new account on Twitter. With these user credentials, we used the 
R package `rtweet` to query the API. Because we work on computers 
with linux operating systems, we made use of linux software 
`crontab` to repeatedly execute R code to submit API queries. Each 
query lasts a user-specified duration. We timed the API queries so 
that there was no time lag between queries. We stored API query 
results in their native JSON format. The R package `rtweet` 
provides functions that parse tweet JSON to R data frames. We then 
conducted all further analyses with the R data frames.

As a proof of principle of our event detection strategy, we chose to 
examine tweets before, during, and after the National Football League's 
2015 Super Bowl. We fitted five latent Dirichlet allocation models. We examined five time periods, each lasting one hour in duration. The first period began approximately 48h before the Super Bowl halftime show. Subsequent time periods started at -24, 0, +24, and +48 hours after the approximate start time of the halftime show.

@@

>>>>

---
title: 'Using tweet content to detect social media events'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "June 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
---

@@

Twitter has changed the way that we communicate. 
In only 280 characters, users speak their piece on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for text 
analysis now enable instructors to use tweets as pedagogical 
tools in mentoring undergraduate research projects.

@@

Our collection of tweets over time served as a valuable resource in our mentoring 
of undergraduate data science research. @nolan2010computing argue for students to 
work with real 
data. 
Working with real data allows students to develop skill not 
only in statistical analysis,
but also in related data science skills, including data 
transfer from online sources, data storage, and using data 
from multiple file formats. Analyzing our collection of tweets gave students 
opportunities to develop skills in all of the above areas.

@@

We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via latent Dirichlet allocation modeling of collections of tweets and 3) reflections on 
using this data set in research mentoring for undergraduate 
students.

@@

### Study design

We sought to validate our hypothesis that we could detect a major social 
media event by examining tweet topic content at distinct time periods. We 
chose the National Football League's Super Bowl as a case study.

As a proof of principle of our event detection strategy, we chose to 
examine tweets before, during, and after the National Football League's 
2015 Super Bowl. We fitted latent Dirichlet allocation models for each of five 
distinct one-hour time periods. The first period began approximately 48h 
before the Super Bowl halftime show. Subsequent time periods started at -24, 
0, +24, and +48 hours after the approximate start time of the halftime show.

@@

### Collecting tweets over time

We include here instructions for creating a tweet collection. First, we created a 
new account on Twitter. With these user credentials, we used the 
R package `rtweet` to query the API. Because we work on computers 
with linux operating systems, we use the linux software 
`crontab` to repeatedly execute R code to submit API queries. Each 
query lasts a user-specified duration. We time the API queries so 
that there is no time lag between queries. We store API query 
results in their native JSON format. The R package `rtweet` 
provides functions that parse tweet JSON to R data frames. We then 
conducted all further analyses with the R data frames.

### Querying Twitter API to get complete tweets

We queried a tweets database, created with the methods described above, to get ID numbers for 
tweets from the desired time periods. We then submitted API queries to Twitter
to get the full content of the tweets, including the tweet text.
We provide below the R code that we used to query the Twitter API to obtain full tweet content.

```{r, eval = FALSE}
rtweet::lookup_tweets()
```

### Tweet structure

Tweets are available as Javascript Object Notation (JSON) objects. Every tweet consists of multiple named fields, each of which is a key-value pair. The number of fields per tweet depends on user settings, retweet status, and other factors.

**PLACE TWEET JSON HERE**

### Parsing text of tweets

We used functions from the `rtweet` package to parse tweet JSON into a data frame. From there, we used `tidytext` R package functions to break the tweet text into individual words for latent Dirichlet allocation. We discarded commonly used "stop words" and emojis.

Latent Dirichlet allocation models require that the corpus be inputted as a 
document by term matrix. Each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document by 
term matrix with the R functions `` from the `` R package.
